---
title: "Reproducible Project-2"
author: "Mohamed"
date: "22/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Reproducible Research Course Project 2  



##Analysis of the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database  


Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

In this report,effect of weather events on personal as well as property damages was studied. Barplots were plotted seperately for the top 8 weather events that causes highest fatalities and highest injuries. Results indicate that most Fatalities and injuries were caused by Tornados.Also, barplots were plotted for the top 8 weather events that causes the highest property damage and crop damage.


####There are two goals of this analysis:
  
- identify the weather events that are most harmful with respect to population health
- identify the weather events that have the greatest economic consequences.    


####Based on our analysis, we conclude that TORNADOS and FLOODS are most harmful weather events in the USA in terms of the risk to human health and economic impact.  



##Data Processing  

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

Storm Data [47Mb] There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation

National Climatic Data Center Storm Events FAQ

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

2.2 Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer the following basic questions about severe weather events.

1.Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2.Across the United States, which types of events have the greatest economic consequences?

2.3 Modus Operandi / Process
2.3.1 Loading the data
The data was downloaded from the above mentioned website and saved on local computer. Then it was loaded on the R using the following code.


####The data source is in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. It is possible to download the source file from the course web site: [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)


```{r, echo=TRUE, results='hide'}
# reading data
Raw_data <- read.csv("repdata_data_StormData.csv", header=TRUE, sep=",")
```


####1.	According to NOAA, the data recording start from Jan. 1950. At that time, they recorded only one event type - tornado. They added more events gradually, and only from Jan 1996 they started recording all events type. Since our objective is comparing the effects of different weather events, we need only to include events that started not earlier than Jan 1996.  

2.3.2 Extracting the required data

This dataset consists of lot of information most of which is not required for our present study. So, here is the code to extract the required data for health and economic impact analysis against weather.

```{r, echo=TRUE, results='hide'}
# subsetting by date
Main_data <- Raw_data
Main_data$BGN_DATE <- strptime(Raw_data$BGN_DATE, "%m/%d/%Y %H:%M:%S")
Main_data <- subset(Main_data, BGN_DATE > "1995-12-31")
```



  

####Namely: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.  

####Therefore, we can limit our data to these variables.  

```{r, echo=TRUE, results='hide'}
Main_data <- subset(Main_data, select = c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```



####Contents of data now are as follows:  
  
EVTYPE – type of event  
FATALITIES – number of fatalities  
INJURIES – number of injuries  
PROPDMG – the size of property damage  
PROPDMGEXP - the exponent values for 'PROPDMG' (property damage)  
CROPDMG - the size of crop damage  
CROPDMGEXP - the exponent values for 'CROPDMG' (crop damage)  



####3.	There are almost 1000 unique event types in EVTYPE column. Therefore, it is better to limit database to a reasonable number. We can make it by capitalizing all letters in EVTYPE column as well as subsetting only non-zero data regarding our target numbers.  


2.3.4 Finding crop damage
Crop damage exponents for each level was listed out and assigned those values for the crop exponent data. Invalid data was excluded by assigning the value as ‘0’. Then crop damage value was calculated by multiplying the crop damage and crop exponent value.The code for this process was listed below

```{r, echo=TRUE, results='hide'}
#cleaning event types names
Main_data$EVTYPE <- toupper(Main_data$EVTYPE)
# eliminating zero data
Main_data <- Main_data[Main_data$FATALITIES !=0 | 
                       Main_data$INJURIES !=0 | 
                       Main_data$PROPDMG !=0 | 
                       Main_data$CROPDMG !=0, ]
```
  
####Now we have 186 unique event types and it seems like something to work with. 
  

##Population health data processing  


####We aggregate fatalities and injuries numbers in order to identify TOP-10 events contributing the total people loss:


```{r, echo = TRUE, results='asis'}
Health_data <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data = Main_data, FUN=sum)
Health_data$PEOPLE_LOSS <- Health_data$FATALITIES + Health_data$INJURIES
Health_data <- Health_data[order(Health_data$PEOPLE_LOSS, decreasing = TRUE), ]
Top10_events_people <- Health_data[1:10,]
knitr::kable(Top10_events_people, format = "markdown")
```



##Economic consequences data processing  


####The number/letter in the exponent value columns (PROPDMGEXP and CROPDMGEXP) represents the power of ten (10^The number). It means that the total size of damage is the product of PROPDMG and CROPDMG and figure 10 in the power corresponding to exponent value.  



####Exponent values are:  
- numbers from one to ten  
- letters  (B or b = Billion, M or m = Million, K or k = Thousand, H or h = Hundred)  
- and symbols "-", "+" and "?" which refers to less than, greater than and low certainty. We have the option to ignore these three symbols altogether.  


####We transform letters and symbols to numbers:

```{r, echo=TRUE, results = 'hide'}
Main_data$PROPDMGEXP <- gsub("[Hh]", "2", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("[Kk]", "3", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("[Mm]", "6", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("[Bb]", "9", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("\\+", "1", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("\\?|\\-|\\ ", "0",  Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- as.numeric(Main_data$PROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Hh]", "2", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Kk]", "3", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Mm]", "6", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Bb]", "9", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("\\+", "1", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("\\-|\\?|\\ ", "0", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- as.numeric(Main_data$CROPDMGEXP)
Main_data$PROPDMGEXP[is.na(Main_data$PROPDMGEXP)] <- 0
Main_data$CROPDMGEXP[is.na(Main_data$CROPDMGEXP)] <- 0
```
  
  


####At last, we create new values of total property damage and total crop damage for analysis (we need ‘dplr’ package for that).  

```{r, echo=TRUE, results = 'hide', message = FALSE, warning = FALSE}
#creating total damage values
library(dplyr)
Main_data <- mutate(Main_data, 
                    PROPDMGTOTAL = PROPDMG * (10 ^ PROPDMGEXP), 
                    CROPDMGTOTAL = CROPDMG * (10 ^ CROPDMGEXP))
```
  
  

####Now we aggregate property and crop damage numbers in order to identify TOP-10 events contributing the total economic loss:

```{r, echo=TRUE, results = 'asis'}
Economic_data <- aggregate(cbind(PROPDMGTOTAL, CROPDMGTOTAL) ~ EVTYPE, data = Main_data, FUN=sum)
Economic_data$ECONOMIC_LOSS <- Economic_data$PROPDMGTOTAL + Economic_data$CROPDMGTOTAL
Economic_data <- Economic_data[order(Economic_data$ECONOMIC_LOSS, decreasing = TRUE), ]
Top10_events_economy <- Economic_data[1:10,]
knitr::kable(Top10_events_economy, format = "markdown")
```



##Results  
  

2.3.5 Finding totals of each incident by event type.

It was observed that " most harmful to population health" events are fatalities and injuries.So,only those events with fatalities and injuries were selecetd.

It was observed that " most harmful to econamic problem“” events are Property and crop damages.So,only those events with property and crop damage were selecetd.

Then for each incident (Fatalities,Injuries, Property damage and Crop damage), the total values were estimated. Code for which is as follow
  

```{r, echo=TRUE, results='asis', message = FALSE, warning = FALSE}
#plotting health loss
library(ggplot2)
g <- ggplot(data = Top10_events_people, aes(x = reorder(EVTYPE, PEOPLE_LOSS), y = PEOPLE_LOSS))
g <- g + geom_bar(stat = "identity", colour = "black")
g <- g + labs(title = "Total people loss in USA by weather events in 1996-2011")
g <- g + theme(plot.title = element_text(hjust = 0.5))
g <- g + labs(y = "Number of fatalities and injuries", x = "Event Type")
g <- g + coord_flip()
print(g)
```
  
  

####Analyzing economic impact on the graph one can conclude that FLOOD, HURRICANE/TYPHOON and STORM SURGE are the main contributors to severe economic consequences out of all event types of weather events. 

```{r, echo=TRUE, results='asis'}
#plotting economic loss
g <- ggplot(data = Top10_events_economy, aes(x = reorder(EVTYPE, ECONOMIC_LOSS), y = ECONOMIC_LOSS))
g <- g + geom_bar(stat = "identity", colour = "black")
g <- g + labs(title = "Total economic loss in USA by weather events in 1996-2011")
g <- g + theme(plot.title = element_text(hjust = 0.5))
g <- g + labs(y = "Size of property and crop loss", x = "Event Type")
g <- g + coord_flip()
print(g)
```